---
title: "COVID-19 Global Dashboard"
author: "Chris Berg"
date: <br>`r format(Sys.Date(), '%d %B %Y') `
output: 
  html_document:
    theme: darkly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(!require(pacman)) install.packages("pacman")
pacman::p_load( tidyverse ,
                ggplot2  ,
                lubridate,
                janitor,
                sf ,
                maps ,
                leaflet, 
                hrbrthemes , 
                plotly )

cases_world = 
  read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv") %>%
  pivot_longer( -c("Province/State":"Long") , 
                names_to = "date" , 
                values_to = "cases" ) %>%
  mutate( date = lubridate::mdy(date) ) %>%
  clean_names()

deaths_world = read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv") %>%
  pivot_longer( -c("Province/State":"Long") , 
                names_to = "date" , 
                values_to = "deaths" ) %>%
  mutate( date = lubridate::mdy(date) ) %>%
  clean_names()

world_daily_df = 
  cases_world %>%
  merge( deaths_world ) %>% 
  mutate(province_state = replace_na(province_state , "Nationwide") ) %>%
  arrange( country_region ,
           desc(date) )

recent_date <- max(world_daily_df$date)
```

The data for this dashboard comes from the [Johns Hopkins University Center for Systems Science and Engineering GitHub repo](https://github.com/CSSEGISandData/COVID-19) which is updated daily. Updates to this page occur around 11.30pm, when JHU updates their data.

## Map of COVID-19, Today

```{r worldmap-data, include=F}

world_today = 
  world_daily_df %>%
  filter( date == recent_date ) %>%
  mutate( cfr = deaths/cases )

world  = 
  st_as_sf(map("world", plot = FALSE, fill = TRUE)) %>% 
  st_transform( crs = 4326 )

proj_crs = st_crs(world)

world_sf = 
  world_today %>% 
  st_as_sf( coords = c("long","lat") ,
            crs = 4326 )

```

Below is a fully-interactive map which highlights the current recorded cases, deaths, and case fatality ratios of COVID-19 (a.k.a. coronavirus or 2019-novel coronavirus) within countries. Click-and-drag to move around the map. Larger red circles indicate a greater number of Coronavirus cases in that country. Clicking on any of the dots will pull-up the country name, along with its headline COVID-19 stats. 

```{r worldmap-int, echo=F , fig.align='center'}

world_sf %>% 
  filter( cases > 0 ) %>%
  st_transform(crs = "epsg:4326") %>%
  leaflet(width = "100%") %>% 
  addProviderTiles(provider = "CartoDB.Positron") %>%
  addCircleMarkers( radius = ~log(cases) , 
                    popup = ~paste0(paste(province_state, country_region , sep = ", ") , 
                                    "<br>", "Confirmed Cases: ", prettyNum(cases, big.mark=",") , 
                                    "<br>", "Deaths: " , prettyNum(deaths, big.mark=",") , 
                                    "<br>", "Case fatality ratio: " , prettyNum(round(cfr,3))),
                    color = 'red' , 
                    stroke = F ) 

```


## COVID-19 infection and mortality trends

```{r data-plot, include=F}

country_df = 
  world_daily_df %>%
  group_by( country_region , date) %>%
  summarise( cases = sum(cases) , 
             deaths = sum(deaths) ) %>%
  mutate( new_cases = cases - dplyr::lag(cases, order_by = date) ,
          new_deaths = deaths - dplyr::lag(deaths, order_by = date) , 
          avg_new_cases = 
            (1/7)*( new_cases +
            dplyr::lag(new_cases, order_by = date, n = 1) + 
            dplyr::lag(new_cases, order_by = date, n = 2) +
            dplyr::lag(new_cases, order_by = date, n = 3) + 
            dplyr::lag(new_cases, order_by = date, n = 4) + 
            dplyr::lag(new_cases, order_by = date, n = 5) + 
            dplyr::lag(new_cases, order_by = date, n = 6)  )
          )

cases_plot = 
  country_df %>%
  ggplot( ) + 
  geom_line( aes( x = date , y = avg_new_cases , color = country_region ) ) + 
  labs( title = "New COVID-19 cases worldwide" , 
        subtitle = "7-day average",
        x = "Date" , 
        y = "New Reported COVID-19 Cases" , 
        color = "Country" ) +
  theme_ft_rc( axis_title_size = 15 )

deaths_plot = 
  country_df %>%
  ggplot( ) + 
  geom_line( aes( x = date , y = deaths , color = country_region ) ) + 
  labs( title = "COVID-19 mortality worldwide" , 
        x = "Date" , 
        y = "Reported COVID-19 Deaths" , 
        color = "Country" ) +
  theme_ft_rc( axis_title_size = 15 )

```


The graphs below show nationwide historical trends in COVID-19 spread, offering additional context to the snapshot above. To pull-up the graph for one specific country, double-click the country name in the legend. Click (once) on any additional countries to compare their trends. New case trends are computed as the [7-day average](https://en.wikipedia.org/wiki/Moving_average#Simple_moving_average) of new cases.

```{r int-plot, echo=FALSE, dpi = 120 }

plotly::ggplotly( cases_plot
                  )

```

```{r int-deaths-plot, echo=FALSE, dpi = 120 }

plotly::ggplotly( deaths_plot
                  )

```